\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-08-19 papers/jss/article.tex
... jss final version, to be submitted.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@310 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameter are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} routine.  }

@@

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses. % response variables are considered.
However, commonly employed estimation procedures
\citep[e.g.,][]{Pol1996}, are not suitable for long time series due to
underflow problems. In contrast, the hidden Markov model is typically
only used for `long' univariate time series \citep[][chapter~1]{Cappe2005}. 
We use the term ``dependent mixture model'' because one of the authors 
(Ingmar Visser) thought it
was time for a new name to relate these models\footnote{Only later did I
find out that \citet{Leroux1992} already coined the term dependent
mixture models in an application with hidden Markov mixtures of
Poisson count data.}.

The fundamental assumption of a dependent mixture model is that at any time 
point, the observations are distributed as a mixture with $n$ components 
(or states), and that time-dependencies between the observations are due to 
time-dependencies between the mixture components (i.e., transition probabilities
between the components). These latter dependencies are assumed to
follow a first-order Markov process. In the models we are considering here, the 
mixture distributions, the initial mixture probabilities and transition 
probabilities can all dependent on covariates $\vc{z}_t$.

@@

In a dependent mixture model, the joint likelihood of observations $\vc{O}_{1:T}$ and 
latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$, given model parameters 
$\greekvec{\theta}$ and covariates $\vc{z}_{1:T} = (\vc{z}_1,\ldots,\vc{z}_T)$, 
can be written as
\begin{equation}
	\Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\greekv{\theta},\vc{z}_{1:T}) =  
	\pi_{i}(\vc{z}_1) \vc{b}_{S_t}(\vc{O}_1|\vc{z}_{1})
	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{S_t}(\vc{O}_{t+1}|\vc{z}_{t+1}),
\end{equation}
where we have the following elements:
\begin{enumerate}

	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
	of $n$ latent classes or states. %; we write for short $S_{i}$ to
	%denote $S_{t}=i$.

@@

	\item $\vc{b}_{S_t}$ is a vector of observation densities
	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
	conditional densities of observations $O_{t}^k$ associated with
	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
	$k=1, \ldots, m$. 
\end{enumerate}

@@

For the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli distribution for the
accuracy variable.  In the models we are considering here, both the
transition probability functions $a_{ij}$ and the initial state
probability functions $\greekv{\pi}$ may depend on covariates as well
as the response distributions $b_{j}^{k}$.  See for example
\citet{Fruhwirth2006} for an overview of hidden Markov models with
extensions.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM algorithm, 
parameters are estimated by iteratively maximising the 
expected joint likelihood of the parameters given the observations and states. 
Let $\greekv{\theta} = (\greekv{\theta}_1, \greekv{\theta}_2,\greekv{\theta}_3)$
be the general parameter vector consisting of three subvectors with parameters 
for the prior model, transition model, and response model respectively. The 
joint log-likelihood can be written as
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $O_{1:T}$. The expected 
log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
forward-backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
\pkg{nnet} package \citep{Venables2002}, and maximisation for 
$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
observations $O^k_t$.

@@

\subsection{Example data: speed}

Throughout this article a data set called \code{speed} is used.  As already 
indicated in the Introduction, it
consists of three time series with three variables: response time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding.  The participant in this experiment
switches between fast responding at chance level and relatively slower
responding at a high level of accuracy.  Interesting hypotheses to
test are: is the switching regime symmetric?  Is there evidence for
two states or does one state suffice?  Is the guessing state actually
a guessing state, i.e., is the probability of a correct response at
chance level (0.5)?

@@

A dependent mixture model is defined by the number of states and the initial 
state, state transition, and response distribution functions. A default 
dependent mixture model can be created with the \code{depmix}-function as 
follows:

@@

The so-defined models needs to be \code{fit}ted with the following
line of code:
\begin{CodeChunk}
\begin{CodeInput}
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log-likelihood and the AIC and BIC
values:
\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34175 (df=7)
AIC:  182.6835 
BIC:  211.275 
\end{CodeOutput}
\end{CodeChunk}
These statistics can also be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
     [,1]      [,2]
[1,]    0 -11.25688
Probalities at zero values of the covariates.
0.999987 1.291798e-05

Transition model for state (component) 1 
Model of type multinomial, formula: ~1
Coefficients: 
[1]  0.000000 -2.392455
Probalities at zero values of the covariates.
0.9162501 0.08374986

Transition model for state (component) 2 
Model of type multinomial, formula: ~1
Coefficients: 
[1] 0.000000 2.139255
Probalities at zero values of the covariates.
0.1053396 0.8946604

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.385492
sd  0.2439376

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.511151
sd  0.1926063 
\end{CodeOutput}
\end{CodeChunk}

@@

Applying \code{summary} to the \code{fit}ted object gives (only transition models
printed here): 
\begin{CodeChunk}
\begin{CodeOutput}
...
Transition model for state (component) 1 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
     [,1]       [,2]
[1,]    0 -0.9215182
[2,]    0  1.8649734
Probalities at zero values of the covariates.
0.7153513 0.2846487

Transition model for state (component) 2 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
     [,1]     [,2]
[1,]    0 2.471442
[2,]    0 3.570856
Probalities at zero values of the covariates.
0.07788458 0.9221154
...
\end{CodeOutput}
\end{CodeChunk}
The summary provides all parameters of the model, also the 
(redundant) zeroes for the base-line category in the multinomial model. 
The summary also prints the transition probabilities
at the zero value of the covariate. Note that scaling of the covariate 
is useful in this regard as it makes interpretation of these intercept probabilities 
easier.

@@

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the Gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
...
Response model(s) for state 1

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.52169
sd  0.2028857

Response model for response 2 
Model of type multinomial, formula: corr ~ 1
Coefficients: 
     [,1]      [,2]
[1,]    0 0.1030554
Probalities at zero values of the covariates.
0.4742589 0.5257411

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.39369
sd  0.2373650

Response model for response 2 
Model of type multinomial, formula: corr ~ 1
Coefficients: 
	   [,1]     [,2]
[1,]    0 2.245514
Probalities at zero values of the covariates.
0.09573715 0.9042629 	
\end{CodeOutput}
\end{CodeChunk}
As can be seen, state 1 has fast response times around chance level, 
whereas state 2 corresponds with slower responding at higher accuracy 
levels.

Note that by specifying multivariate observations in terms of a list, the 
variables are considered conditionally independent (given the states). 
Conditionally \emph{dependent} variables must be handled as a single element in
the list. Effectively, this means specifying a multivariate response model. The 
only multivariate response model currently implemented in \pkg{depmix} is for
multivariate normal variables.

@@

Similarly to the transition matrix, covariates on the prior 
probabilities of the latent states (or classes in this case), are 
defined by using a one-sided formula: 
\begin{CodeChunk}
\begin{CodeInput}
> balance$age <- balance$age-5
> set.seed(1)
> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
    family=list(multinomial(), multinomial(), multinomial(),
    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
    prior=~age, initdata=balance)
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Mixture probabilities model 
Model of type multinomial, formula: ~age
     [,1]       [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
...
\end{CodeOutput}
\end{CodeChunk}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

The \pkg{depmixS4} package was designed with the aim of making it relatively
easy to add new response distributions (as well as possibly new prior and 
transition models). 
To make this possible, the EM routine simply calls the \code{fit} methods 
of the separate response models without needing to know how they work. As a 
consequence, adding user-specified response models is straightforward. 
User-defined distributions should extend the \code{response}-class 
and have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters, e.g., the standard deviation in the Gaussian 
	response model
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

In the \code{speed} data example, it may be more appropriate to model
the response times with an exgaus rather than Gaussian distribution.  
To do so, we first define an exgaus-class extending the response-class:
\begin{CodeChunk}
\begin{CodeInput}
setClass("exgaus", contains="response")
\end{CodeInput}
\end{CodeChunk}

@@

Only the constructor and the fit method are provided here; the complete code 
can be found in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss} and \pkg{gamlss.distr} 
packages \citep{Stasinopoulos2009a,Stasinopoulos2009b} for 
computing the \code{dens}ity and for \code{fit}ting the parameters.

The constructor method is defined as: 
\begin{CodeChunk}
\begin{CodeInput}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

@@

We first create all the response models that we want as a double list: 
\begin{CodeChunk}
\begin{CodeInput}
rModels <- list(
  list(
    exgaus(rt,pstart=c(5,.1,.1)),
    GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
  ),
  list(
    exgaus(rt,pstart=c(6,.1,.1)),
    GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
  )
)
\end{CodeInput}
\end{CodeChunk}
Next, we define the transition and prior probability models using the 
\code{transInit} function (which produces a transInit model, which also extends 
the response class): 
\begin{CodeChunk}
\begin{CodeInput}
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
\end{CodeInput}
\end{CodeChunk}
Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
\begin{CodeChunk}
\begin{CodeInput}
> mod <- makeDepmix(response=rModels,transition=transition,
    prior=inMod,stat=FALSE)
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

Using \code{summary} will print the fitted parameters. Note that the 
use of \code{makeDepmix} allows the possibility of, say, fitting a 
Gaussian in one state and an exgaus distribution in another state.

>>>>

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} routine.}

@@

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses.  However,
commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
not suitable for long time series due to underflow problems.  In
contrast, the hidden Markov model is typically only used for `long'
univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
term ``dependent mixture model'' because one of the authors (Ingmar
Visser) thought it was time for a new name to relate these
models\footnote{Only later did I find out that \citet{Leroux1992}
already coined the term dependent mixture models in an application
with hidden Markov mixtures of Poisson count data.}.

The fundamental assumption of a dependent mixture model is that at any
time point, the observations are distributed as a mixture with $n$
components (or states), and that time-dependencies between the
observations are due to time-dependencies between the mixture
components (i.e., transition probabilities between the components).
These latter dependencies are assumed to follow a first-order Markov
process.  In the models we are considering here, the mixture
distributions, the initial mixture probabilities and transition
probabilities can all depend on covariates $\vc{z}_t$.

@@

In a dependent mixture model, the joint likelihood of observations
$\vc{O}_{1:T}$ and latent states $\vc{S}_{1:T} = (S_1,\ldots,S_T)$,
given model parameters $\greekv{\theta}$ and covariates $\vc{z}_{1:T}
= (\vc{z}_1,\ldots,\vc{z}_T)$, can be written as:
\begin{equation}
	\Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\greekv{\theta},\vc{z}_{1:T}) =  
	\pi_{i}(\vc{z}_1) \vc{b}_{S_t}(\vc{O}_1|\vc{z}_{1})
	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{S_t}(\vc{O}_{t+1}|\vc{z}_{t+1}),
\end{equation}
where we have the following elements:
\begin{enumerate}

	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
	of $n$ latent classes or states.

@@

	\item $\vc{b}_{S_t}$ is a vector of observation densities
	$b_{j}^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that
	provide the conditional densities of observations $O_{t}^k$
	associated with latent class/state $j$ and covariate $\vc{z}_t$,
	$j=1, \ldots, n$, $k=1, \ldots, m$.

@@

For the example data above, $b_j^k$ could be a Gaussian distribution
function for the response time variable, and a Bernoulli distribution
for the accuracy variable.  In the models we are considering here,
both the transition probability functions $a_{ij}$ and the initial
state probability functions $\greekv{\pi}$ may depend on covariates as
well as the response distributions $b_{j}^{k}$.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximising the
expected joint likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response model respectively.  The joint
log-likelihood can be written as
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. The expected 
log likelihood 
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
\end{equation}
can be written as
%\begin{equation}
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\ln \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
%\end{equation}
where the expected values $\xi_t(j,k) =  P(S_t = k, S_{t-1} = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) = P(S_t = j|\vc{O}_{1:T},
\vc{z}_{1:T},\greekv{\theta}')$ can be computed effectively by the 
forward-backward algorithm \citep[see e.g.,][]{Rabiner1989}. The Maximisation 
step consists of the maximisation of (\ref{eq:Q}) for $\greekv{\theta}$. As the 
right hand side of (\ref{eq:Q}) consists of three separate parts, we can 
maximise separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and 
$\greekv{\theta}_3$. In common models, maximisation for $\greekv{\theta}_1$ and 
$\greekv{\theta}_2$ is performed by the \code{nnet.default} routine in the 
\pkg{nnet} package \citep{Venables2002}, and maximisation for 
$\greekv{\theta}_3$ by the standard \code{glm} routine. Note that for the latter 
maximisation, the expected values $\gamma_t(j)$ are used as prior weights of the 
observations $O^k_t$.

@@

\subsection[Example data: speed]{Example data: \code{speed}}

Throughout this article a data set called \code{speed} is used.  As
already indicated in the Introduction, it consists of three time
series with three variables: response time, accuracy, and a covariate
Pacc which defines the relative pay-off for speeded versus accurate
responding.  The participant in this experiment switches between fast
responding at chance level and relatively slower responding at a high
level of accuracy.  Interesting hypotheses to test are: is the
switching regime symmetric?  Is there evidence for two states or does
one state suffice?  Is the guessing state actually a guessing state,
i.e., is the probability of a correct response at chance level (0.5)?

@@

A dependent mixture model is defined by the number of states and the
initial state, state transition, and response distribution functions.
A dependent mixture model can be created with the
\code{depmix}-function as follows:

@@

The so-defined model needs to be \code{fit}ted with the following
line of code:
\begin{CodeChunk}
\begin{CodeInput}
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log-likelihood and the AIC and BIC
values:
\begin{CodeChunk}
\begin{CodeInput}
> fm
\end{CodeInput}
\begin{CodeOutput}
Convergence info: Log likelihood converged to within tol. 
'log Lik.' -84.34 (df=7)
AIC:  182.68 
BIC:  211.28 
\end{CodeOutput}
\end{CodeChunk}
These statistics can also be extracted using \code{logLik}, \code{AIC}
and \code{BIC}, respectively.  By comparison, a 1-state model for
these data, i.e. assuming there is no mixture, has a log-likelihood of
-305.33, and 614.66, and 622.83 for the AIC and BIC respectively.
Hence, the 2-state model fits the data much better than the 1-state
model.  Note that the 1-state model can be specified using \code{mod <-
depmix(rt~1, data=speed, nstates=1)}, although this model is trivial
as it will simply return the mean and the sd of the rt variable.

@@

\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Initial state probabilties model 
Model of type multinomial, formula: ~1
Coefficients: 
     [,1]      [,2]
[1,]    0 -11.25
Probalities at zero values of the covariates.
0.999 1.292e-05

Transition model for state (component) 1 
Model of type multinomial, formula: ~1
Coefficients: 
[1]  0.000 -2.392
Probalities at zero values of the covariates.
0.9163 0.0837

Transition model for state (component) 2 
Model of type multinomial, formula: ~1
Coefficients: 
[1] 0.000 2.139
Probalities at zero values of the covariates.
0.105 0.895

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.385
sd  0.244

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.511
sd  0.193 
\end{CodeOutput}
\end{CodeChunk}

@@

Applying \code{summary} to the \code{fit}ted object gives (only transition models
printed here): 
\begin{CodeChunk}
\begin{CodeOutput}
...
Transition model for state (component) 1 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
     [,1]       [,2]
[1,]    0 -0.9215
[2,]    0  1.865
Probalities at zero values of the covariates.
0.7154 0.2846

Transition model for state (component) 2 
Model of type multinomial, formula: ~scale(Pacc)
Coefficients: 
     [,1]     [,2]
[1,]    0 2.471
[2,]    0 3.571
Probalities at zero values of the covariates.
0.0779 0.9221
...
\end{CodeOutput}
\end{CodeChunk}
The summary provides all parameters of the model, also the (redundant)
zeroes for the base-line category in the multinomial model.  The
summary also prints the transition probabilities at the zero value of
the covariate.  Note that scaling of the covariate is useful in this
regard as it makes interpretation of these intercept probabilities
easier.

@@

Multivariate data can be modelled by providing a list of formulae as
well as a list of family objects for the distributions of the various
responses.  In above examples we have only used the response times
which were modelled with the Gaussian distribution.  The accuracy
variable in the \code{speed} data can be modelled with a multinomial
by specifying the following:
\begin{CodeChunk}
\begin{CodeInput}
> set.seed(1)
> mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
    family=list(gaussian(),multinomial()),
    transition=~scale(Pacc),instart=runif(2))
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
This provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
...
Response model(s) for state 1

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 5.522
sd  0.2029

Response model for response 2 
Model of type multinomial, formula: corr ~ 1
Coefficients: 
     [,1]      [,2]
[1,]    0 0.1031
Probalities at zero values of the covariates.
0.4743 0.5257

@@

Response model for response 1 
Model of type gaussian, formula: rt ~ 1
Coefficients: 
[1] 6.394
sd  0.2374

Response model for response 2 
Model of type multinomial, formula: corr ~ 1
Coefficients: 
	   [,1]     [,2]
[1,]    0 2.2455
Probalities at zero values of the covariates.
0.0957 0.9043 	
\end{CodeOutput}
\end{CodeChunk}
As can be seen, state 1 has fast response times and accuracy is
approximately at chance level (.474), whereas state 2 corresponds with
slower responding at higher accuracy levels (.904).

Note that by specifying multivariate observations in terms of a list,
the variables are considered conditionally independent (given the
states).  Conditionally \emph{dependent} variables must be handled as
a single element in the list.  Effectively, this means specifying a
multivariate response model.  The only multivariate response model
currently implemented in \pkg{depmixS4} is for multivariate normal
variables.

@@

Similarly to the transition matrix, covariates on the prior
probabilities of the latent states (or classes in this case), are
defined by using a one-sided formula:
\begin{CodeChunk}
\begin{CodeInput}
> balance$age <- balance$age-5
> set.seed(1)
> mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
    family=list(multinomial(), multinomial(), multinomial(),
    multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
    prior=~age, initdata=balance)
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}
Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.  More formally,
\code{depmix} models extend the class of \code{mix} models by adding
the transition models.  As for fitting \code{mix} models: as can be
seen in equation~\ref{eq:Q}, the EM algorithm can be applied by simply
dropping the second summand containing the transition parameters, and 
this is implemented as such in the EM algorithms in \pkg{depmixS4}.

The summary of the \code{fit}ted model gives the following (only the
prior model is shown here):
\begin{CodeChunk}
\begin{CodeInput}
> summary(fm)
\end{CodeInput}
\begin{CodeOutput}
Mixture probabilities model 
Model of type multinomial, formula: ~age
     [,1]       [,2]
[1,]    0 -2.518
[2,]    0  0.551
Probalities at zero values of the covariates.
0.9254 0.0746
...
\end{CodeOutput}
\end{CodeChunk}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

The \pkg{depmixS4} package was designed with the aim of making it
relatively easy to add new response distributions (as well as possibly
new prior and transition models).  To make this possible, the EM
routine simply calls the \code{fit} methods of the separate response
models without needing access to the internal workings of these
routines.  Referring to equation~\ref{eq:Q}, the EM algorithm calls
separate fit functions for each part of the model, the prior
probability model, the transition models, and the response models.  As
a consequence, adding user-specified response models is
straightforward.

User-defined distributions should extend the \code{response}-class and
have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters, e.g., the standard deviation in the Gaussian 
	response model
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

In the \code{speed} data example, it may be more appropriate to model
the response times with an exgaus rather than a Gaussian distribution.
To do so, we first define an exgaus-class extending the
response-class:
\begin{CodeChunk}
\begin{CodeInput}
setClass("exgaus", contains="response")
\end{CodeInput}
\end{CodeChunk}

@@

Only the constructor and the fit methods are provided here; the
complete code can be found in the help file of the \code{makeDepmix}
function.  The \code{exgaus} example uses the \pkg{gamlss} and
\pkg{gamlss.distr} packages
\citep{Stasinopoulos2009a,Stasinopoulos2009b} for computing the
\code{dens}ity and for \code{fit}ting the parameters.

The constructor method return an object of class \code{exgaus}, and is
defined as follows:
\begin{CodeChunk}
\begin{CodeInput}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
  standardGeneric("exgaus"))

@@

We first create all the response models that we need as a double list: 
\begin{CodeChunk}
\begin{CodeInput}
rModels <- list(
  list(
    exgaus(rt,pstart=c(5,.1,.1)),
    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
  ),
  list(
    exgaus(rt,pstart=c(6,.1,.1)),
    GLMresponse(corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
  )
)
\end{CodeInput}
\end{CodeChunk}
Next, we define the transition and prior probability models using the 
\code{transInit} function (which produces a transInit model, which also extends 
the response class): 
\begin{CodeChunk}
\begin{CodeInput}
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nst=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
\end{CodeInput}
\end{CodeChunk}
Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
\begin{CodeChunk}
\begin{CodeInput}
> mod <- makeDepmix(response=rModels,transition=transition,
    prior=inMod,stat=FALSE)
> fm <- fit(mod)
\end{CodeInput}
\end{CodeChunk}

Using \code{summary} will print the fitted parameters.  Note that the
use of \code{makeDepmix} allows the possibility of, say, fitting a
Gaussian in one state and an exgaus distribution in another state.
Note also that according to the AIC and BIC, the model with the exgaus
describes the data much better than the same model in which the
response times are modeled as gaussian.
